---
title: "p8105_hw2_mcm2335"
author: "Madeleine McGuinness"
date: "2023-10-03"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(readxl)
```

## Set Up
```{r}
#set working directory
setwd("~/Desktop/R_Building_Blocks_P8105/p8105_hw2_mcm2335")
```

## Question 1 

## Import CSV Data Tables
```{r}
pols_month = read_csv(file = "./data/pols-month.csv")

unemployment = read_csv(file = "./data/unemployment.csv")

snp = read_csv(file = "./data/snp.csv")
```

## Clean Data Table: Pols-month
```{r}
names(pols_month) # names look clean

pols_month_2 <- pols_month |>
  separate(col = mon, into=c('year', 'month', 'day', sep='-')) #sep mon column

pols_month_2 <- pols_month_2 |>
  select(-'-') #remove '-' column

pols_month_2 <- pols_month_2 |>
  mutate(month = if_else(month == "01", "January", if_else(month == "02", "February", if_else(month == "03", "March", if_else(month == "04", "April", if_else(month == "05", "May", if_else(month == "06", "June", if_else(month == "07", "July", if_else(month == "08", "August", if_else(month == "09", "September", if_else(month == "10", "October", if_else(month == "11", "November", "December")))))))))))) #mutate month numbers to month names

pols_month_2 <- pols_month_2 |>
  mutate(president = if_else(prez_dem == 1, "dem", if_else(prez_gop == 1, "gop", NA_character_))) # new president value using prez_dem and prez_gop

pols_month_3 <- pols_month_2 |>
  select(-prez_dem, -prez_gop) # remove prez_dem and prez_gop variables

pols_month_final <- pols_month_3 |>
  select(-day) # remove day variable
```

## Clean Data Table: snp
```{r}
names(snp) # names look clean

snp_2 <- snp |>
  separate(col = date, into=c('month', 'day', 'year', sep='/')) #sep date column

snp_2 <- snp_2 |>
  select(-'/') #remove '/' column

snp_2 <- snp_2 |>
  mutate(month = if_else(month == "1", "January", if_else(month == "2", "February", if_else(month == "3", "March", if_else(month == "4", "April", if_else(month == "5", "May", if_else(month == "6", "June", if_else(month == "7", "July", if_else(month == "8", "August", if_else(month == "9", "September", if_else(month == "10", "October", if_else(month == "11", "November", "December")))))))))))) #mutate month numbers to month names

snp_final <- snp_2 |>
  select(year, month, day, close) # arrange final snp dataset by year, month

snp_final <- snp_final |>
  transform(year = as.numeric(year)) # tranform year to numeric 

snp_final <- snp_final |>
  mutate(year = if_else(year < 16, year + 2000, year + 1900)) |>
  select(-day) |>
  transform(year = as.character(year))# mutate year to 4 digits instead of 2 digits for consistency with pols_month_final dataset and remove day, year back to character value
```

## Clean Data Table: unemployment
```{r}
names(unemployment) #need to clean to make names all lowercase

unemployment_2 <- janitor::clean_names(unemployment) #cleaned names

unemployment_3 <- unemployment_2 |>
  pivot_longer(
    jan:dec, 
    names_to = "month",
    values_to = "unemployment") #pivoted from wide to long format

unemployment_final <- unemployment_3 |>
  mutate(month = if_else(month == "jan", "January", if_else(month == "feb", "February", if_else(month == "mar", "March", if_else(month == "apr", "April", if_else(month == "may", "May", if_else(month == "jun", "June", if_else(month == "jul", "July", if_else(month == "aug", "August", if_else(month == "sep", "September", if_else(month == "oct", "October", if_else(month == "nov", "November", "December")))))))))))) |>
  transform(year = as.character(year))#mutate month abbreviations to month names, transform year to character for consistency
```

## Merge Data Tables
```{r}
combined_data =
  left_join(pols_month_final, snp_final)

combined_data_final =
  left_join(combined_data, unemployment_final)
```

## Answer 
Note: I did not use inline R code for this answer, as I am unable to knit when using inline R code. 

The 'pols_month_final' dataset has 822 observations and 9 variables and indicates the political affiliation of presidents by month and year from 1947 to 2015 and the distribution of governors and senators by political affiliation. The 'snp_final' dataset has 787 observations and 3 variables and has 'close' information by month and year from 1950 to 2015. The 'unemployment_final' dataset has 816 observations and 3 variables and provides national unemployment rates by month and year from 1948 to 2015. The 'combined_data_final' dataset merges all of these variables by month and year and has 822 observations, 11 variables, and 53 total NAs.

## Question 2 

## Import Excel Data Table, Tidy, Mutate New Variables

```{r}
mr_trash_wheel = readxl::read_excel("./data/202309 Trash Wheel Collection Data.xlsx", sheet ='Mr. Trash Wheel', range = cell_rows(2:586)) |>
  mutate(source = "mr_trash_wheel") |>
  janitor::clean_names() |>
  transform(year = as.numeric(year)) |>
  select(-x15, -x16, -sports_balls) #excluded non-data rows by only reading in 2:586, removed non-dumspter variables (x15,x16, sports_balls), mutated source variable for merging later on, cleaned names

mr_trash_wheel <- mr_trash_wheel |>
  mutate(homes_powered_new = (weight_tons*500)/30)

professor_trash_wheel = readxl::read_excel("./data/202309 Trash Wheel Collection Data.xlsx", sheet ='Professor Trash Wheel', range = cell_rows(2:108)) |>
  mutate(source = "prof_trash_wheel") |>
  janitor::clean_names()#excluded non-data rows by only reading in 2:108, mutated source variable for merging later on, cleaned names

professor_trash_wheel <- professor_trash_wheel |>
  mutate(homes_powered_new = (weight_tons*500)/30)
  
gwynnda_trash_wheel = readxl::read_excel("./data/202309 Trash Wheel Collection Data.xlsx", sheet ='Gwynnda Trash Wheel', range = cell_rows(2:157)) |>
  mutate(source = "gwynnda_trash_wheel") |>
  janitor::clean_names()#excluded non-data rows by only reading in 2:157, removed non-dumspter variables (x15,x16, sports_balls), mutated source variable for merging later on, cleaned names. There are two rows for dumpster 21, and thus there are 155 obs but 154 dumpsters

gwynnda_trash_wheel <- gwynnda_trash_wheel |>
  mutate(homes_powered_new = (weight_tons*500)/30)

```

## Merging Data Sets

```{r}
trash_wheel_combined = bind_rows(mr_trash_wheel, professor_trash_wheel, gwynnda_trash_wheel) #combined all datasets by common columns
```

## Analysis of Data Sets
```{r}
weight <- trash_wheel_combined |>
  group_by(source) |>
  summarise(sum(weight_tons)) # total weight of trash from professor trash wheel = 216 tons

cigbutts <- trash_wheel_combined |>
  filter(source == "gwynnda_trash_wheel" & year == 2021 & month == "July")|>
  summarise(sum(cigarette_butts)) # total cig butts collected in July of 2021 by Gwynnda was 16,300 cig butts

```

## Question 2 Answer
Note: I did not use inline R code for this answer, as I am unable to knit when using inline R code.

The mr_trash_wheel dataset has 584 observations of 15 variables. The professor_trash_wheel dataset has 106 observations and 15 variables. The gwynnda_trash_wheel dataset has 155 observations and 14 variables. Each dataset had variables 'dumpster,' 'month,' 'year,' 'date,' 'weight_tons' (representing weight of trash collected in tons), 'volumbe_cubic_yards,' (of trash collected), 'plastic_bottles' (collected in trash), 'polystyrene' (collected in trash), 'cigarette_buts' (collected in trash), 'glass bottles' (collected in trash), 'plastic_bags' (collected in trash), 'wrappers' (collected in trash), and 'homes_powered', a variable representing how many homes could be powered in a day by the trash collected if each ton generated 500 kilowatts of energy and each home required 30 kilowatts per day. As some dumpsters were missing this calculation, I mutated a new variable 'homes_powered_new' using this equation: (weight_tons*500)/30. Finally, I mutated a source variable for each dataset representing where the data came from for merging. Only the gwynnda_trash_wheel dataset did not have the variable 'glass_bottles', and thus every observation from the gwynnda source in the merged dataset has an NA value for this variable. The final merged dataset has 845 observations and 15 variables.

As for the analysis, the total weight of trash collected by Professor Trash Wheel was 216 tons. The total count of cigarette butts collected by Gwynnda in July 2021 was 16,300 cigarette butts.

## Question 3

## Import  CSVs 
```{r}
mci_baseline = read_csv(file = "./data/MCI_baseline.csv", skip = 1) |> #originally 483 participants
  janitor::clean_names() |>
  transform(id = as.character(id)) |>
  mutate(sex = if_else(sex == 0, "Female", if_else(sex == 1, "Male", NA_character_))) |>
  mutate(apoe4 = if_else(apoe4 == 0, "Non-Carrier", if_else(apoe4 == 1, "Carrier", NA_character_))) |>
  filter(age_at_onset != ".")# skipped first row which was metadata #97 participants after filtering out individuals without MCI at follow up


mci_amyloid = read_csv(file = "./data/mci_amyloid.csv", skip = 1) |> #487 participants originally
  janitor::clean_names() |>
  rename(id = study_id) |>
  transform(id = as.character(id))
# skipped first row which was metadata
```
## Merge Datasets
```{r}
mci_combined <- inner_join(mci_baseline, mci_amyloid, by = "id")
```

## Save CSV to data folder
```{r}
write.csv(mci_combined, file = "/Users/robynrehak/Desktop/R_Building_Blocks_P8105/p8105_hw2_mcm2335/mci_combined.csv", row.names = FALSE)
```

## Analysis
```{r}
#97/483 #20.08% developed MCI

avg_age_baseline <- mean(mci_baseline$current_age) #65.61

female <- mci_baseline |>
  filter(sex == "Female")

table(female$apoe4) #46 in total, #30 carriers, #16 non carriers
#30/46 #65.22%
#16/46 #34.78%
```
## Question 3 Answer
Note: I did not use inline R code for this answer, as I am unable to knit when using inline R code.

When importing both my mci_baseline and mci_amyloid datasets, it was important to add the parameter in my 'read_csv' function to skip the first row of both datasets, as this row contained metadata in the csv files and not actual observations or column names. After importing the csv files, I cleaned the variable names and converted the 'id' variable to a character vector for consistency across the two datasets to help with merging later on. Then, for the 'mci_baseline' dataset, I used the metadata of the csv to convert the numeric '1' and '0' values to 'Female' and 'Male' character values for the 'sex' variable, and to 'Carrier' or 'Non-carrier' character values respectively for the 'apoe4' variable. To remove individuals without MCI at follow up, I filtered all individuals with the "." value for 'age_at_onset' out of the final dataset for mci_baseline. For mci_amyloid, I renamed the variable 'study_id' to 'id' and converted it to a character value in order to merge the dataset with mci_baseline. I merged the two datasets using the 'inner_join' function with the common 'id' variable to remove any participants that were not included in both datasets. 'Mci_baseline' after tidying has 97 observations and 6 variables, including variables for ID, age at baseline, sex, education level, carrier status for apoe4, and current age while 'Mci_amyloid' after tidying has 487 observations and 6 variables and included information on the biomarker amyloid B 42/40 ratio at baseline, as well as 2, 4, 6, and 8 years after baseline. The final dataset had 11 variables, as 'id' was merged. 

In terms of the analysis, 483 participants were originally included in the mci_baseline dataset, and after excluding individuals free of MCI at follow-up, 97 participants remained, meaning 97 participants developed MCI. After merging mci_baseline and mci_amyloid, only 94 participants remained, meaning that 3 participants with MCI in the mci_baseline dataset did not exist in the mci_amyloid dataset, and that mci_amyloid had 393 participants not in the mci_baseline dataset. 

The average baseline age for individuals who developed MCI in the mci_baseline dataset was 65.61 years. 65.22% of individuals who developed MCI were carriers for apoe4. 

The final combined dataset was saved as a csv in the data folder.

